---
pagetitle: The multiple regression model
output: 
  revealjs::revealjs_presentation:
    incremental: false
    theme: solarized
    self_contained: false
    # reveal_plugins: ["menu","notes","chalkboard"]
    reveal_plugins: ["menu"]
    highlight: pygments
    center: true
    transition: none
    background_transition: none 
    reveal_options:
      # chalkboard:
      #   theme: whiteboard
      #   toggleNotesButton: true
      #   toggleChalkboardButton: true
      menu:
        numbers: true
      slideNumber: true
      previewLinks: false
    fig_caption: true
    pandoc_args:
    - --indented-code-classes
    - lineNumbers
    css: mystyle.css
    
--- 

<section>

<h1>The multiple regression model</h1>

Based on Stock and Watson, ch. 6

<br>

<h2>[Jesper Bagger](mailto:jesper.bagger@rhul.ac.uk)</h2>

<h3>EC2203 | Royal Holloway | 2020/21</h3>

</section>


```{r results='asis', echo=FALSE, include=FALSE}
library(AER) # load Applied Econometrics with R library
library(parameters) # Load parameters library for robust SE computation
library(plotly) # Include library plotly for 3D plots
library(reshape2) # LInclude library for flexible reshaping of dataframes, used in conjunctino w/ plotly for £D-plots
data(CASchools) # Load CASchools data
# Generate a couple of useful variables
CASchools$STR <- CASchools$students/CASchools$teachers  # Student-teacher ratio
CASchools$Score <- (CASchools$read + CASchools$math)/2  # Student test score
```

# Outline

1. Omitted variable bias

2. Holding other variables constant

3. The multiple regression model

4. The OLS estimator in multiple regression

# Omitted variable bias

## The simple regression model

<div class="box">
$$Y_i = \beta_0 + \beta_1 X_i + u_i; \quad i = 1,2,...,n$$

The causal effect $\beta_1$ is well estimated by OLS if

1. Zero conditional mean: $\mathrm{E}(u_i|X_i) = 0$

2. The data point $(X_i,Y_i;i=1,2,...,n)$ are i.i.d. draws from the joint distribution of $X$ and $Y$

3. Large outliers of $X_i$ and $Y_i$ are unlikely

</div>


## Test scores, student-teacher ratio, and percentage of English learners

```{r echo=TRUE, error=FALSE, warning=FALSE, out.width = "70%"}
library(AER) # load Applied Econometrics with R library
library(parameters) # Load parameters library for robust SE computation
data(CASchools) # Load CASchools data
# Generate a couple of useful variables
CASchools$STR <- CASchools$students/CASchools$teachers  # Student-teacher ratio
CASchools$Score <- (CASchools$read + CASchools$math)/2  # Student test score
```

## Test scores and student-teacher ratio

```{r echo=FALSE, error=FALSE, warning=FALSE, out.width = "70%"}
lm1 <- lm(Score ~ STR,data= CASchools)
plot(CASchools$STR,CASchools$Score, 
     col = "blue",
     xlab = "Student-teacher ratio", 
     ylab = "Test score", 
     xlim = c(10, 30), 
     ylim = c(600, 720))
abline(lm1,
       lwd = 3,
       col = "red")
```

## Test scores and percentage of English learners

```{r echo=FALSE, error=FALSE, warning=FALSE, out.width = "70%"}
lm2 <- lm(Score ~ english,data= CASchools)
plot(CASchools$english,CASchools$Score, 
     col = "blue",
     xlab = "Learning English", 
     ylab = "Test score", 
     xlim = c(0, 100), 
     ylim = c(600, 720))
abline(lm2,
       lwd = 3,
       col = "red")
```


## Percentage of English learners and student-teacher ratio

```{r echo=FALSE, error=FALSE, warning=FALSE, out.width = "70%"}
lm3 <- lm(english ~ STR,data= CASchools)
plot(CASchools$STR,CASchools$english, 
     col = "blue",
     xlab = "Student-teacher ratio", 
     ylab = "Learning English", 
     xlim = c(10, 30), 
     ylim = c(0, 100))
abline(lm3,
       lwd = 3,
       col = "red")
```

## Omitted variable bias

```{r echo=FALSE, error=FALSE, warning=FALSE, out.width = "80%"}
par(mfrow=c(1,3))

plot(CASchools$STR,CASchools$Score, 
     col = "blue",
     xlab = "Student-teacher ratio", 
     ylab = "Test score", 
     xlim = c(10, 30), 
     ylim = c(600, 720))
abline(lm1,
       lwd = 3,
       col = "red")

plot(CASchools$english,CASchools$Score, 
     col = "blue",
     xlab = "Learning English", 
     ylab = "Test score", 
     xlim = c(0, 100), 
     ylim = c(600, 720))
abline(lm2,
       lwd = 3,
       col = "red")

plot(CASchools$STR,CASchools$english, 
     col = "blue",
     xlab = "Student-teacher ratio", 
     ylab = "Learning English", 
     xlim = c(10, 30), 
     ylim = c(0, 100))
abline(lm3,
       lwd = 3,
       col = "red")
```

## Omitted variable bias in simple regression

<div class="box"> 
$$Y_i = \beta_0 + \beta_1 X_i + u_i; \quad i = 1,2,...,n$$

**Omitted variable bias** is the bias in the OLS estimator of the causal effect of $X$ on $Y$ that arises when the regressor $X$ is **correlated** with an **omitted variable**

Omitted variable bias arises when two conditions are met:

1. $X$ is **correlated** with the omitted variable

2. The omitted variable is a **determinant** of $Y$
</div>

## Omitted variables and the least squares assumptions

- Because the omitted variable is a determinant of $Y$, it is contained in the error term $u$

- If the omitted variable correlates with the included $X$, 

  $$\mathrm{E}(u|X) \neq 0$$
  
- Omitted variables **violate** the least squares assumptions for causal inference: OLS estimator **biased**, **inconsistent**

  $$\hat{\beta}_1 \overset{p}{\rightarrow} \beta_1 + \frac{\sigma_{Xu}}{\sigma^2_X}$$
  
# Holding other variables constant

## Holding the omitted variable constant

- The omitted variable problem arise because an excluded determinant of $Y$ **correlates** with $X$

  Solution: estimate the effect of $X$ on $Y$ **holding the omitted variable constant**

- Works because constant variables cannot correlate with other variables

## Holding Percentage English learners constant

All districts 

```{r echo=FALSE, error=FALSE, warning=FALSE, out.width = "70%"}
# Scatter plots of STR, Score and english data
par(mfrow=c(1,2)) 
# STR vs Score
plot(CASchools$STR,CASchools$Score,
     xlab = "Student-teacher ratio",
     ylab = "Test score",
     col = "gray",
     xlim = c(10, 30),
     ylim = c(600, 720))
abline(lm1,
       lwd = 3,
       col = "gray")
# STR vs english
plot(CASchools$STR,CASchools$english,
     col = "gray",
     xlab = "Student-teacher ratio",
     ylab = "Learning English",
     xlim = c(10, 30),
     ylim = c(0, 100))
abline(lm3,
       lwd = 3,
       col = "gray")
```

## Holding Percentage English learners constant

Districts where percentage learning English $\leq$ 1.9%

```{r echo=FALSE, error=FALSE, warning=FALSE, out.width = "80%"}
# Select dataframe subset with based on english
CASchools1 <- subset(CASchools, english < 1.9)
CASchools2 <- subset(CASchools, english >= 1.9 & english < 8.8)
CASchools3 <- subset(CASchools, english >= 8.8 & english < 23)
CASchools4 <- subset(CASchools, english >= 23)
# Run regression of Score onto STR in each subsample
lm1.1 <- lm(Score ~ STR, data = CASchools1) # Score onto STR in CASchools1
lm1.2 <- lm(Score ~ STR, data = CASchools2) # Score onto STR in CASchools2
lm1.3 <- lm(Score ~ STR, data = CASchools3) # Score onto STR in CASchools3
lm1.4 <- lm(Score ~ STR, data = CASchools4) # Score onto STR in CASchools4
# Run regression of english onto STR in each subsample
lm3.1 <- lm(english ~ STR, data = CASchools1) # english onto STR in CASchools1
lm3.2 <- lm(english ~ STR, data = CASchools2) # english onto STR in CASchools2
lm3.3 <- lm(english ~ STR, data = CASchools3) # english onto STR in CASchools3
lm3.4 <- lm(english ~ STR, data = CASchools4) # english onto STR in CASchools4
# Scatter plots of STR, Score and english data
par(mfrow=c(1,2)) 
# STR vs Score
plot(CASchools$STR,CASchools$Score,
     xlab = "Student-teacher ratio",
     ylab = "Test score",
     col = "gray",
     xlim = c(10, 30),
     ylim = c(600, 720))
abline(lm1,
       lwd = 3,
       col = "gray")
points(CASchools1$STR,CASchools1$Score,
       col = "blue")
abline(lm1.1,
       lwd = 3,
       col = "blue")
# STR vs english
plot(CASchools$STR,CASchools$english,
     col = "gray",
     xlab = "Student-teacher ratio",
     ylab = "Learning English",
     xlim = c(10, 30),
     ylim = c(0, 100))
abline(lm3,
       lwd = 3,
       col = "gray")
points(CASchools1$STR,CASchools1$english,
       col = "blue")
abline(lm3.1,
       lwd = 3,
       col = "blue")

```

## Holding Percentage English learners constant

Districts where percentage learning English b/w 1.9%-8.8%

```{r echo=FALSE, error=FALSE, warning=FALSE, out.width = "80%"}
# Scatter plots of STR, Score and english data
par(mfrow=c(1,2)) 
# STR vs Score
plot(CASchools$STR,CASchools$Score,
     xlab = "Student-teacher ratio",
     ylab = "Test score",
     col = "gray",
     xlim = c(10, 30),
     ylim = c(600, 720))
abline(lm1,
       lwd = 3,
       col = "gray")
points(CASchools1$STR,CASchools1$Score,
       col = "blue")
abline(lm1.1,
       lwd = 3,
       col = "blue")
points(CASchools2$STR,CASchools2$Score,
       col = "red")
abline(lm1.2,
       lwd = 3,
       col = "red")
# STR vs english
plot(CASchools$STR,CASchools$english,
     col = "gray",
     xlab = "Student-teacher ratio",
     ylab = "Learning English",
     xlim = c(10, 30),
     ylim = c(0, 100))
abline(lm3,
       lwd = 3,
       col = "gray")
points(CASchools1$STR,CASchools1$english,
       col = "blue")
abline(lm3.1,
       lwd = 3,
       col = "blue")
points(CASchools2$STR,CASchools2$english,
       col = "red")
abline(lm3.2,
       lwd = 3,
       col = "red")

```

## Holding Percentage English learners constant

Districts where percentage learning English b/w 8.8%-23%

```{r echo=FALSE, error=FALSE, warning=FALSE, out.width = "80%"}
# Scatter plots of STR, Score and english data
par(mfrow=c(1,2)) 
# STR vs Score
plot(CASchools$STR,CASchools$Score,
     xlab = "Student-teacher ratio",
     ylab = "Test score",
     col = "gray",
     xlim = c(10, 30),
     ylim = c(600, 720))
abline(lm1,
       lwd = 3,
       col = "gray")
points(CASchools1$STR,CASchools1$Score,
       col = "blue")
abline(lm1.1,
       lwd = 3,
       col = "blue")
points(CASchools2$STR,CASchools2$Score,
       col = "red")
abline(lm1.2,
       lwd = 3,
       col = "red")
points(CASchools3$STR,CASchools3$Score,
       col = "orange")
abline(lm1.3,
       lwd = 3,
       col = "orange")
# STR vs english
plot(CASchools$STR,CASchools$english,
     col = "gray",
     xlab = "Student-teacher ratio",
     ylab = "Learning English",
     xlim = c(10, 30),
     ylim = c(0, 100))
abline(lm3,
       lwd = 3,
       col = "gray")
points(CASchools1$STR,CASchools1$english,
       col = "blue")
abline(lm3.1,
       lwd = 3,
       col = "blue")
points(CASchools2$STR,CASchools2$english,
       col = "red")
abline(lm3.2,
       lwd = 3,
       col = "red")
points(CASchools3$STR,CASchools3$english,
       col = "orange")
abline(lm3.3,
       lwd = 3,
       col = "orange")

```

## Holding Percentage English learners constant

Districts where percentage learning English $>$ 23%

```{r echo=FALSE, error=FALSE, warning=FALSE, out.width = "80%"}
# Scatter plots of STR, Score and english data
par(mfrow=c(1,2)) 
# STR vs Score
plot(CASchools$STR,CASchools$Score,
     xlab = "Student-teacher ratio",
     ylab = "Test score",
     col = "gray",
     xlim = c(10, 30),
     ylim = c(600, 720))
abline(lm1,
       lwd = 3,
       col = "gray")
points(CASchools1$STR,CASchools1$Score,
       col = "blue")
abline(lm1.1,
       lwd = 3,
       col = "blue")
points(CASchools2$STR,CASchools2$Score,
       col = "red")
abline(lm1.2,
       lwd = 3,
       col = "red")
points(CASchools3$STR,CASchools3$Score,
       col = "orange")
abline(lm1.3,
       lwd = 3,
       col = "orange")
points(CASchools4$STR,CASchools4$Score,
       col = "green")
abline(lm1.4,
       lwd = 3,
       col = "green")
# STR vs english
plot(CASchools$STR,CASchools$english,
     col = "gray",
     xlab = "Student-teacher ratio",
     ylab = "Learning English",
     xlim = c(10, 30),
     ylim = c(0, 100))
abline(lm3,
       lwd = 3,
       col = "gray")
points(CASchools1$STR,CASchools1$english,
       col = "blue")
abline(lm3.1,
       lwd = 3,
       col = "blue")
points(CASchools2$STR,CASchools2$english,
       col = "red")
abline(lm3.2,
       lwd = 3,
       col = "red")
points(CASchools3$STR,CASchools3$english,
       col = "orange")
abline(lm3.3,
       lwd = 3,
       col = "orange")
points(CASchools4$STR,CASchools4$english,
       col = "green")
abline(lm3.4,
       lwd = 3,
       col = "green")

```

## Holding Percentage English learners constant (cont'd)

$$Score_i = \beta_0 + \beta_1 STR_i + u_i; \quad i=1,2,...,n$$

|                                                             | $\hat{\beta}_0$              | $\hat{\beta}_1$            | 
| ---------------                                             |--:                           | --:                        | 
| All districts                                               | $\underset{(10.36)}{698.93}$ | $-\underset{(0.52)}{2.28}$ | 
| Districts where percentage learning English $\leq$ 1.9%     | $\underset{(19.03)}{650.24}$ | $\underset{(0.98)}{0.77}$  | 
| Districts where percentage learning English b/w 1.9%-8.8%   | $\underset{(16.90)}{700.75}$ | $-\underset{(0.82)}{1.87}$ | 
| Districts where percentage learning English b/w 8.8%-23%    | $\underset{(18.16)}{695.87}$ | $-\underset{(0.92)}{2.20}$ | 
| Districts where percentage learning English $>$ 23%         | $\underset{(15.57)}{653.07}$ | $-\underset{(0.76)}{0.87}$ | 

## Multiple regression analysis

- Dividing the data into groups is a flexible way of addressing omitted variable bias; 

- The approach also has disadvantages:

  - How to split data based on continuous omitted variable?
  
  - Impractible with many omitted variables
  
  - Does not yield a single estimate of the effect of $X$ on $Y$ holding a set of other variables constant
  
- **Multiple regression analysis** is a way of circumventing these disadvantages

# The multiple regression model

## The population regression function with two regressors

- Let $Y$ be the random variable we want to model as a function of **two** random variables $X_1$ and $X_2$

  We say $Y$ is the **dependent variable**, and $X_1$ and $X_2$ are the **independent variables** or **regressors**

- Consider the linear **population regression function** 

  <div class="box">
  $$\mathrm{E}(Y|X_1,X_2) = \beta_0 + \beta_1 X_1 + \beta_2 X_2$$
  </div>
  
  The population regression function is the relationship b/w $Y$ and $X_1$ and $X_2$ that holds **on average** in the population
  
<!-- - The population regression function has **population regression coefficients** $\beta_0$, $\beta_1$ and $\beta_2$ -->

## Interpreting the slope coefficients 

- The population regression function has **population regression coefficients** $\beta_0$, $\beta_1$ and $\beta_2$

- Consider predictions for two observations with the **same** values of $X_2$, but values of $X_1$ that differ by $\Delta X_1$; then,

  $$\beta_1 = \frac{\mathrm{E}(Y|X_1+\Delta X_1,X_2)  - \mathrm{E}(Y|X_1,X_2)}{\Delta X_1}$$
  
- $\beta_1$ is difference in predicted $Y$ b/w obs. w/ unit difference in $X_1$, **holding $X_2$ constant**, or **controlling for $X_2$**

  We sometimes say $\beta_1$ is the **partial effect** on $Y$ of $X_1$. Analogously, $\beta_2$ is the **partial effect** on $Y$ of $X_2$

## The error term

<!-- - Let $u$ be the **error term**: the error made when predicting $Y$ by its conditional mean; that is, $u \equiv Y - \mathrm{E}(Y|X_1,X_2)$ -->

- Let $u$ be the **error term**: the error made when predicting $Y$ by its conditional mean $\mathrm{E}(Y|X_1,X_2)$; that is,

  <div class="box">
  $$u \equiv Y - \mathrm{E}(Y|X_1,X_2)$$
  </div>

- Defines $Y$ as a linear function of $X_1$ and $X_2$, and $u$:
  
  <div class="box">
  $$Y = \beta_0 + \beta_1 X_1 +  \beta_2 X_2 + u$$
  </div>

- The error term $u$ contains **all** determinants of the dependent variable $Y$ that are **not** $X_1$ nor $X_2$

## The population regression model with two regressors

- Estimation problem: learn about $\beta_0$, $\beta_1$, and $\beta_2$ from **random sample** of data from the population

  $$(Y_i,X_{1i},X_{2i}; i=1,...,n)$$

- Each observation $i$ in the sample follows the population regression model; that is,

  <div class="box">
  $$Y_i = \beta_0 + \beta_1 X_{1i} + \beta_1 X_{2i} + u_i; \quad i=1,...,n$$
  </div>
  
- The multiple regression model generalizes easily to encompass $k$ regressors

# The OLS estimator in multiple regression 

## The data

```{r echo=FALSE, error=FALSE, warning=FALSE, out.width = "100%"}
fig <- plot_ly(x = CASchools$STR, y = CASchools$english, z = CASchools$Score, 
               marker = list(color = CASchools$Score, colorscale = list(c(0, 1), c("lightblue", "blue")), showscale = FALSE))
fig <- fig %>% add_markers()
fig <- fig %>% layout(scene = list(
                     xaxis = list(title = 'Student-teacher ratio'),
                     yaxis = list(title = 'Percentage English learners'),
                     zaxis = list(title = 'Test score')))
fig
```

## The best plane  

- The OLS estimator finds regression coefficients that puts the fitted regression plane as **close to** the data as possible

- The distance between the regression plane is measured by the **sum of squared in-sample prediction errors**

  With two regressors, OLS estimators $\hat{\beta}_0,\hat{\beta}_1,\hat{\beta}_2$ solve

  <div class="box">
  $$\min_{\hat{\beta}_0,\hat{\beta}_1,\hat{\beta}_2} \sum_{i=1}^n \big(\underset{ \text{in-sample prediction error} }{\underbrace{Y_i - \hat{\beta}_0 - \hat{\beta}_1 X_{1i} - \hat{\beta}_2 X_{2i} }} \big)^2$$
  </div>

## OLS terminology revisited

- The **sample regression function** is 

  $$\hat{\beta}_0 + \hat{\beta}_1 X_1 + \hat{\beta}_2 X_2$$

- The **predicted value** of $Y_i$ given $X_{1i}$ and $X_{2i}$ is 

  $$\hat{Y}_i = \hat{\beta}_0 + \hat{\beta}_1 X_{1i} + \hat{\beta}_2 X_{2i}$$

- The **residual** for the $i$th observation 

  $$\hat{u}_i = Y_i - \hat{Y}_i = Y_i - \hat{\beta}_0 - \hat{\beta}_1 X_{1i} - \hat{\beta}_2 X_{2i}$$

## Implementing the OLS estimator in R

```{r echo=TRUE, error=FALSE, warning=FALSE, out.width = "70%"}

lm4 <- lm(Score ~ STR + english, data = CASchools)
summary(lm4)
```

## ... with robust standard errors

```{r echo=TRUE, error=FALSE, warning=FALSE, out.width = "70%"}
parameters(lm4, robust = TRUE, vcov_type = "HC1")
```

## The estimated regression plane

```{r echo=FALSE, error=FALSE, warning=FALSE, out.width = "100%"}
lm4.resSq <- lm4$residuals^2

#Graph Resolution (more important for more complex shapes)
graph_reso <- 1

# Setup Axis
axis_x <- seq(min(CASchools$STR), max(CASchools$STR), by = graph_reso)
axis_y <- seq(min(CASchools$english), max(CASchools$english), by = graph_reso)

# Sample points
lm4.surface <- expand.grid(STR = axis_x, english = axis_y, KEEP.OUT.ATTRS = FALSE)
lm4.surface$Score <- predict.lm(lm4, newdata = lm4.surface)
lm4.surface <- acast(lm4.surface, english ~ STR, value.var = "Score") #y ~ x


fig <- plot_ly(x = CASchools$STR, y = CASchools$english, z = CASchools$Score,
               marker = list(color = lm4.resSq, colorscale = list(c(0, 1), c("lightblue", "blue")), showscale = FALSE, showlegend = FALSE))
fig <- fig %>% add_markers()
fig <- fig %>% layout(scene = list(
                     xaxis = list(title = 'Student-teacher ratio'),
                     yaxis = list(title = 'Percentage English learners'),
                     zaxis = list(title = 'Test score')))
fig <- add_trace(p = fig,
                       z = lm4.surface,
                       x = axis_x,
                       y = axis_y,
                       type = "surface",
                       colorscale = list(c(0, 1), c("pink", "red")), showscale = FALSE,showlegend = FALSE, opacity = 0.8) 
fig
```

# Summary

## Summary

- Omitted variable bias arises if an omitted regressor (i) correlates w/ included regressors, (ii) is a determinant of $Y$

- The multiple regression model is a linear regression model w/ multiple regressors $X_1,...,X_k$. Associated w/ each regressor is a coefficient $\beta_1,...,\beta_k$

- Coefficient $\beta_1$ measures the expected change in $Y$ from a one-unit change in $X_1$, holding all the other regressors constant (with analogous interpretations of $\beta_2,..,\beta_k$)

- The OLS estimator in the multiple regression model minimizes the sum of squared residuals, as in the simple linear regression model
